Cubical Types with Positive and Negative Polarities

### 2014-05-19 ring.tex
separate syntax from denotation
fixed several typos

Regular first-order types are viewed as 1-dimensional cubes. Generalize to
$n$-dimensional cubes.

@@

For cubical types, their denotations, usually written as $\cubt$, are
$n$-dimensional cubical sets with polarities. An $n$-dimensional cubical set
$\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and a
negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
cubical set of a lower dimension; the 1-dimensional case reduces to
$\nodet{S}{~}$ which is essentially the denotation of conventional
first-order types. As we explain below, cubical sets are closed under sums,
differences, products.

Examples of 1-dimensional cubical sets include all the sets of the form
$\nodet{S}{~}$ where $S$ is finite set denoting a regular first-order
types $t$. A 2-dimensional cubical set looks like $\nodet{S_1}{S_2}$ which
intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
denotations are $S_1$ and $S_2$ respectively. A 3-dimensional cubical type
looks like $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
corresponds to the iterated difference of the appropriate types
$(t_1-t_2)-(t_3-t_4)$ where the successive colors from the root encode the
sign as follows: white in the outermost box corresponds to a positive
polarity and gray in the outermost box corresponds to a negative polarity;
then, as we move inwards the polarities alternate for every change of color.

We now give the denotation of cubical types:
\[\begin{array}{rcl}
\denc{t} &=& \nodet{\den{t}}{~} = 
  \nodet{\nodet{\den{t}}{~}}{~} = \ldots \\
\\
\denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
\denc{\tau_1 * \tau_2} &=& \denc{\tau_1} \otimes \denc{\tau_2} \\
\denc{- \tau} &=& \ominus \denc{\tau}
\end{array}\]
where the operations $\oplus$, $\otimes$, and $\ominus$ are defined below.
It is possible for cubical types not be balanced in general but it is always
possible to embed an $n$-dimensional set into a larger dimension using the
equivalence in the first clause which formalizes that $t$ is equivalent to
$t-0$ and to $(t-0)-0$ etc.  In the recursive definitions below, we assume
that the types have been extended to be balanced as needed:
\[\begin{array}{rcl}
\ominus~S &=& S \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
\\
S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
S_1 \otimes S_2 &=& S_1 \times S_2 \\
S \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{S \otimes \cubt_1}{S \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt}
\end{array}\]

\begin{figure*}
\[\begin{array}{c}
\nodet{S_1}{S_2}
\quad\otimes\quad
\nodet{(\nodet{S_3}{S_4})}{(\nodet{S_5}{S_6})} = 
\\
\nodet{(\nodet{{(\nodet{S_1 \times S_3}{S_1 \times S_4})}}
              {{(\nodet{S_1 \times S_5}{S_1 \times S_6})}})}
      {(\nodet{{(\nodet{S_2 \times S_3}{S_2 \times S_4})}}
              {{(\nodet{S_2 \times S_5}{S_2 \times S_6})}})}
\end{array}\]
\caption{\label{mult}Example of multiplication of two cubical types.}
\end{figure*}

The definition of negation is natural: it recursively swaps the positive and
negative subspaces. We note that the general definitions of sum and product
reduce to the conventional sums and products for 1-dimensional types. Also
the sum of 2-dimensional types reduces to the sum used in the \textbf{Int}
construction. The interesting new development is the product definition. The
example in Fig.~\ref{mult} shows that in general the product of an
$n$-dimensional type with an $m$-dimensional type gives an
$(n+m)$-dimensional type. For low dimensions such as the ones used in the
example above, it is possible to visualize what is happening in coordinate
space (see Fig.~\ref{multv}).

\begin{figure*}
\begin{tikzpicture}
\node[left] at (0,0) {$S_1^+$};
\node[right] at (0.6,0) {$S_2^-$};
\draw[-] (0,0) -- (0.6,0);
\node at (1.6,0) {$\otimes$}; 
%%
\node[below left] at (2.5,-0.5) {$S_3^+$};
\node[below right] at (3.5,-0.5) {$S_4^-$};
\draw[-] (2.5,-0.5) -- (3.5,-0.5);
\draw[-] (2.5,-0.5) -- (2.5,0.5);
\node[above left] at (2.5,0.5) {$S_6^-$};
\node[above right] at (3.5,0.5) {$S_5^+$};
\draw[-] (2.5,0.5) -- (3.5,0.5);
\draw[-] (3.5,-0.5) -- (3.5,0.5);
%% 
\node at (4.5,0) {$=$};
%% 
\node[left] at (7.5,0.75) {$(S_2 \times S_3)^-$};
\node[right] at (9.5,0.75) {$(S_2 \times S_4)^+$};
\node[above right] at (10.2,1.2) {$(S_2 \times S_5)^-$};
\node[above left] at (8.2,1.2) {$(S_2 \times S_6)^+$};
%%
\node[left] at (7.5,-0.75) {$(S_1 \times S_3)^+$};
\node[right] at (9.5,-0.75) {$(S_1 \times S_4)^-$};
\node[above right] at (10.2,-0.3) {$(S_1 \times S_5)^+$};
\node[above] at (8.2,-0.3) {$(S_1 \times S_6)^-$};
%%
\draw[-] (7.5,0.75) -- (9.5,0.75);
\draw[-] (9.5,0.75) -- (10.2,1.2);
\draw[-] (10.2,1.2) -- (8.2,1.2);
\draw[-] (8.2,1.2) -- (7.5,0.75);
%%
\draw[-] (7.5,-0.75) -- (9.5,-0.75);
\draw[-] (9.5,-0.75) -- (10.2,-0.3);
\draw[-,dashed] (10.2,-0.3) -- (8.2,-0.3);
\draw[-,dashed] (8.2,-0.3) -- (7.5,-0.75);
%%
\draw[-] (7.5,0.75) -- (7.5,-0.75);
\draw[-] (9.5,0.75) -- (9.5,-0.75);
\draw[-] (10.2,1.2) -- (10.2,-0.3);
\draw[-,dashed] (8.2,1.2) -- (8.2,-0.3);
\end{tikzpicture}
\caption{\label{multv}Visualization of multiplication of two cubical types.}
\end{figure*}

>>>>

Regular first-order types are viewed as 0-dimensional cubes. Generalize to
$n$-dimensional cubes.

@@

For cubical types, their denotations, usually written as $\cubt$, are
$n$-dimensional cubical sets with polarities. An $n$-dimensional cubical set
$\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and a
negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
cubical set of a lower dimension; the 0-dimensional case reduces to the
denotation of conventional first-order types. As we explain below, cubical
sets are closed under sums, differences, products.

Examples of 0-dimensional cubical sets include all the finite sets denoting a
regular first-order type. A 1-dimensional cubical set looks like
$\nodet{S_1}{S_2}$ which intuitively corresponds to the difference $t_1 -
t_2$ of the two types whose denotations are $S_1$ and $S_2$ respectively. A
2-dimensional cubical type looks like
$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
corresponds to the iterated difference of the appropriate types
$(t_1-t_2)-(t_3-t_4)$ where the product of the successive signs from the
outermost box encodes the sign.

We now give the denotation of cubical types:
\[\begin{array}{rcl}
\denc{t} &=& \den{t} \\
\denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
\denc{\tau_1 * \tau_2} &=& \denc{\tau_1} \otimes \denc{\tau_2} \\
\denc{- \tau} &=& \ominus \denc{\tau}
\end{array}\]
where the operations $\oplus$, $\otimes$, and $\ominus$ are defined below.
It is possible for cubical types not be balanced in general but it is always
possible to embed an $n$-dimensional set into a larger dimension using the
fact (to be justified later) that $t$ is equivalent to $t-0$ and to $(t-0)-0$
and so on, i.e.,
\[
\den{t} = \nodet{\den{t}}{\phantom{\den{t}}} = 
  \nodet{\nodet{\den{t}}{\phantom{\den{t}}}}
        {\phantom{{\nodet{\den{t}}{\phantom{\den{t}}}}}} = \ldots \\
\]
In the recursive definitions below, we assume that the types have been
extended to be balanced as needed:
\[\begin{array}{rcl}
\ominus~S &=& S \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
\\
S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
S_1 \otimes S_2 &=& S_1 \times S_2 \\
S \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{S \otimes \cubt_1}{S \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt}
\end{array}\]

\begin{figure*}
\[\begin{array}{c}
\nodet{S_1}{S_2}
\quad\otimes\quad
\nodet{(\nodet{S_3}{S_4})}{(\nodet{S_5}{S_6})} \quad= \\
\\
\nodet{(\nodet{{(\nodet{S_1 \times S_3}{S_1 \times S_4})}}
              {{(\nodet{S_1 \times S_5}{S_1 \times S_6})}})}
      {(\nodet{{(\nodet{S_2 \times S_3}{S_2 \times S_4})}}
              {{(\nodet{S_2 \times S_5}{S_2 \times S_6})}})}
\end{array}\]
\caption{\label{mult}Example of multiplication of two cubical sets.}
\end{figure*}

\noindent The definition of negation is natural: it recursively swaps the
positive and negative subspaces. We note that the general definitions of sum
and product reduce to the conventional sums and products for 0-dimensional
types. Also the sum of 1-dimensional types reduces to the sum used in the
\textbf{Int} construction and that addition and negation do not increase the
dimensions. The interesting new development is the product definition. The
example in Fig.~\ref{mult} illustrates a general pattern: the product of an
$n$-dimensional type with an $m$-dimensional type gives an
$(n+m)$-dimensional type. For low dimensions such as the ones used in the
example above, it is possible to visualize what is happening in coordinate
space (see Fig.~\ref{multv}).

\begin{figure*}
\begin{tikzpicture}
\node[left] at (0,0) {$S_1^+$};
\node[right] at (0.6,0) {$S_2^-$};
\draw[-] (0,0) -- (0.6,0);
\node at (1.6,0) {$\otimes$}; 
%%
\node[below left] at (2.5,-0.5) {$S_3^+$};
\node[below right] at (3.5,-0.5) {$S_4^-$};
\draw[-] (2.5,-0.5) -- (3.5,-0.5);
\draw[-] (2.5,-0.5) -- (2.5,0.5);
\node[above left] at (2.5,0.5) {$S_5^-$};
\node[above right] at (3.5,0.5) {$S_6^+$};
\draw[-] (2.5,0.5) -- (3.5,0.5);
\draw[-] (3.5,-0.5) -- (3.5,0.5);
%% 
\node at (5,0) {$=$};
%% 
\node[left] at (7.5,0.75) {$(S_2 \times S_3)^-$};
\node[right] at (9.5,0.75) {$(S_2 \times S_4)^+$};
\node[above right] at (10.2,1.2) {$(S_2 \times S_6)^-$};
\node[above left] at (8.2,1.2) {$(S_2 \times S_5)^+$};
%%
\node[left] at (7.5,-0.75) {$(S_1 \times S_3)^+$};
\node[right] at (9.5,-0.75) {$(S_1 \times S_4)^-$};
\node[above right] at (10.2,-0.3) {$(S_1 \times S_6)^+$};
\node[above] at (8.2,-0.3) {$(S_1 \times S_5)^-$};
%%
\draw[-] (7.5,0.75) -- (9.5,0.75);
\draw[-] (9.5,0.75) -- (10.2,1.2);
\draw[-] (10.2,1.2) -- (8.2,1.2);
\draw[-] (8.2,1.2) -- (7.5,0.75);
%%
\draw[-] (7.5,-0.75) -- (9.5,-0.75);
\draw[-] (9.5,-0.75) -- (10.2,-0.3);
\draw[-,dashed] (10.2,-0.3) -- (8.2,-0.3);
\draw[-,dashed] (8.2,-0.3) -- (7.5,-0.75);
%%
\draw[-] (7.5,0.75) -- (7.5,-0.75);
\draw[-] (9.5,0.75) -- (9.5,-0.75);
\draw[-] (10.2,1.2) -- (10.2,-0.3);
\draw[-,dashed] (8.2,1.2) -- (8.2,-0.3);
\end{tikzpicture}
\caption{\label{multv}Visualization of multiplication of two cubical sets.}
\end{figure*}
